---
title: "Comparison"
format: html
editor_options: 
  chunk_output_type: console
---

```{r}
if(!rlang::is_installed('cubar')){
  devtools::install_github("mt1022/cubar", dependencies = TRUE)
}

suppressPackageStartupMessages(library(Biostrings))
library(cubar)
library(tidyverse)
library(GGally)
library(ggbreak)
```

### Generate sequence files required for calculating indexes
```{r}
yeast_cds_qc <- check_cds(yeast_cds)
Biostrings::writeXStringSet(yeast_cds_qc, 'yeast_cds_qc.fasta')
heg <- head(yeast_exp[order(-yeast_exp$fpkm), ], n = 500)
Biostrings::writeXStringSet(yeast_cds[names(yeast_cds) %in% heg$gene_id], 'yeast_heg.fasta')
```

### CodonW (Version: 1.4.4)

Generate fop.coa required for calculation of Fop in CodonW
```{bash}
codonw yeast_cds_qc.fasta -coa_cu -nomenu -silent  -c_type 2 -f_type 4 -nowarn -all_indices
```

Generate cai.coa required for calculation of CAI in CodonW
```{r}
cf_all <- count_codons(yeast_cds)
heg <- head(yeast_exp[order(-yeast_exp$fpkm), ], n = 500)
cf_heg <- cf_all[heg$gene_id, ]
rscu_heg <- est_rscu(cf_heg)
cai_order <- c('UUU', 'UCU','UAU','UGU','UUC','UCC','UAC','UGC','UUA','UCA','UAA','UGA','UUG','UCG','UAG','UGG','CUU','CCU','CAU','CGU','CUC','CCC','CAC','CGC','CUA','CCA','CAA','CGA','CUG','CCG','CAG','CGG','AUU','ACU','AAU','AGU','AUC','ACC','AAC','AGC','AUA','ACA','AAA','AGA','AUG','ACG','AAG','AGG','GUU','GCU','GAU','GGU','GUC','GCC','GAC','GGC','GUA','GCA','GAA','GGA','GUG','GCG','GAG','GGG')
cai_order <- gsub('U', 'T', cai_order)
cai_order <- left_join(data.frame(cai_order), rscu_heg, join_by('cai_order' == 'codon'))
cai_order$w_cai[is.na(cai_order$w_cai)] <- 1
write.table(cai_order$w_cai, "cai.coa", col.names = F, row.names = F, quote = F)
```

Record running time of CodonW
```{bash}
# CAI
for i in {1..5}; do time codonw yeast_cds_qc.fasta "yeast_cai${i}.txt" "yeast_cai${i}.blk" -cai_file cai.coa -c_type 2 -nomenu -cai -nowarn; done

# Fop
for i in {1..5}; do time codonw yeast_cds_qc.fasta "yeast_fop${i}.txt" "yeast_fop${i}.blk" -fop_file fop.coa -f_type 4 -nomenu -fop -nowarn; done

# ENC
for i in {1..5}; do time codonw yeast_cds_qc.fasta "yeast_enc${i}.txt" "yeast_enc${i}.blk"  -nomenu -enc -nowarn -silent; done

# GC
for i in {1..5}; do time codonw yeast_cds_qc.fasta "yeast_gc${i}.txt" "yeast_gc${i}.blk"  -nomenu -gc -nowarn -silent; done

# GC3s
for i in {1..5}; do time codonw yeast_cds_qc.fasta "yeast_gc3s${i}.txt" "yeast_gc3s${i}.blk"  -nomenu -gc3s -nowarn -silent; done

# RSCU
for i in {1..5}; do time codonw yeast_cds_qc.fasta "yeast_rscu${i}.txt" "yeast_rscu${i}.blk"  -nomenu -cutot -nowarn -silent; done
```

### cubar

Record running time of cubar
```{r}
# CAI
cf_all <- count_codons(yeast_cds)
heg <- head(yeast_exp[order(-yeast_exp$fpkm), ], n = 500)
cf_heg <- cf_all[heg$gene_id, ]
rscu_heg <- est_rscu(cf_heg)
in_fa <- 'yeast_cds_qc.fasta'

for(i in 1:5){
    print(system.time(
    cai <- get_cai(count_codons(Biostrings::readDNAStringSet(in_fa)), rscu_heg)
    ))
}

# ENC
for(i in 1:5){
    print(system.time(
    enc <- get_enc(count_codons(Biostrings::readDNAStringSet(in_fa)))
    ))
}

# RSCU
for(i in 1:5){
    print(system.time(
    rscu <- est_rscu(count_codons(Biostrings::readDNAStringSet(in_fa)))
    ))
}

# GC
for(i in 1:5){
    print(system.time(
    gc <- get_gc(count_codons(Biostrings::readDNAStringSet(in_fa)))
    ))
}

# GC3s
for(i in 1:5){
    print(system.time(
    gc3s <- get_gc3s(count_codons(Biostrings::readDNAStringSet(in_fa)))
    ))
}

# Fop
# optimal codons from results of CodonW
op_codonw <- c('UCU','UGU','UUC','UCC','UAC','UUG','CGU','CAC','CCA','CAA','AUU','ACU','AUC','ACC','AAC','AGA','AAG','GUU','GCU','GGU','GUC','GCC','GAC','GAA')
op_codonw <- gsub('U', 'T', op_codonw) 
for(i in 1:5){
    print(system.time(
    fop <- get_fop(count_codons(Biostrings::readDNAStringSet(in_fa)),  op = op_codonw)
    ))
}

# tAI
for(i in 1:5){
    print(system.time(
    tai <- get_tai(count_codons(Biostrings::readDNAStringSet(in_fa)), est_trna_weight(yeast_trna_gcn))
    ))
}
```

### CodonU (Version: 1.1.2)

Record running time of CodonU
```{python}
from CodonU import analyzer as an
from Bio.SeqIO import parse
import time
```

```{python}
# CAI
# There was an error generated when running these codes.
ref_fa = 'yeast_heg.fasta'
sequences = list(parse(ref_fa, "fasta"))
ref_seqs = [str(seq.seq) for seq in sequences]
in_fa = 'yeast_cds_qc.fasta'

records = parse(in_fa, "fasta")
cai_all = {}
tic = time.time()
for record in records:
    seq_q = str(record.seq)
    cai = an.cai(seq_q, references=ref_seqs, genetic_code=1)
    cai_all[record.id] = cai

toc = time.time()
print((toc-tic))

# ENC
for i in range(5):
    tic = time.time()
    enc = an.calculate_enc(in_fa, genetic_code_num = 1, gene_analysis=True)
    toc = time.time()
    print((toc-tic))
    
# RSCU
for i in range(5):
    tic = time.time()
    rscu = an.calculate_rscu(in_fa, genetic_code_num = 1, min_len_threshold = 0, gene_analysis = False)
    toc = time.time()
    print((toc-tic))
    
# GC
for i in range(5):
    records = parse(in_fa, "fasta")
    gc_all = {}
    tic = time.time()
    for record in records:
        seq_q = str(record.seq)
        gc = an.gc_123(seq_q)
        gc_all[record.id] = gc
    toc = time.time()
    print((toc-tic))
    
# tAI
url = 'https://gtrnadb.ucsc.edu/GtRNAdb2/genomes/eukaryota/Scere3/'
anti_codon_dict = an.get_anticodon_count_dict(url = url, database='GtRNAdb')
# anti_codon_dict:{'IGC': 11, 'GGC': 0, 'CGC': 0, 'UGC': 5, 'ICC': 0, 'GCC': 16, 'CCC': 2, 'UCC': 3, 'IGG': 2, 'GGG': 0, 'CGG': 0, 'UGG': 10, 'IGU': 11, 'GGU': 0, 'CGU': 1, 'UGU': 4, 'IAC': 14, 'GAC': 0, 'CAC': 2, 'UAC': 2, 'IGA': 11, 'GGA': 0, 'CGA': 1, 'UGA': 3, 'GCU': 4, 'ICG': 6, 'GCG': 0, 'CCG': 1, 'UCG': 0, 'CCU': 1, 'UCU': 11, 'IAG': 0, 'GAG': 1, 'CAG': 0, 'UAG': 3, 'CAA': 10, 'UAA': 7, 'GAA': 10, 'GUU': 10, 'CUU': 14, 'UUU': 7, 'GUC': 16, 'CUC': 2, 'UUC': 14, 'GUG': 7, 'CUG': 1, 'UUG': 9, 'IAU': 13, 'GAU': 0, 'nan': 0, 'UAU': 2, 'CAU': 10, 'GUA': 8, 'UCA': 0, 'CUA': 0, 'UUA': 0, 'GCA': 4, 'CCA': 6}

# run 5 times, and delete the generated best_fit.py file after each run.
tic = time.time()
tai_df, abs_wi_df, rel_wi_df = an.calculate_gtai(in_fa, genetic_code_num = 1, anticodon_dict = anti_codon_dict)
toc = time.time()
print((toc-tic))
```

### DAMBE (Verson: 7.3.6)

Manually calculate the running time of the last step (from clicking run to the end of the run)

### Comparison of runtime
The average running time of each tool and each index (tool_comp.txt):
Tool	Time	Index
cubar	0.0346	CAI
cubar	0.0494	ENC
cubar	0.0364	RSCU
cubar	0.0312	GC
cubar	0.0386	GC3s
cubar	0.0380	FOP
cubar	0.0560	tAI
CodonW	0.3018	CAI
CodonW	0.3014	ENC
CodonW	0.2238	RSCU
CodonW	0.3028	GC
CodonW	0.2972	GC3s
CodonW	0.4290	FOP
DAMBE	1.7840	CAI
DAMBE	3.1980	ENC
DAMBE	0.7460	RSCU
CodonU	2.5756	ENC
CodonU	2.1783	RSCU
CodonU	1.8720	GC
CodonU	15.8972	tAI

```{r}
library(ggsci)
df <- read.table('../tool_comp.txt', header = T)
df$Tool <- factor(df$Tool, levels = unique(df$Tool))
df1 <- df %>%
       group_by(Index) %>%
       mutate(relative_time = Time/first(Time)) %>%
       filter(Tool != 'cubar')

ggplot(df1,aes(x=factor(Tool,levels =unique(Tool)), 
               y=relative_time, 
               fill=factor(Index,levels = unique(Index)), 
               ))+
    labs(
      x="",  
      y="Runtime relative to cubar",  
      fill=""
    ) + 
    geom_bar(
        stat="identity",
        position=position_dodge(0.5),
        width=0.5
    ) + 
    scale_y_log10() +
    scale_fill_nejm()+
    theme_bw() +
    theme(panel.grid=element_blank(),
       # plot.title = element_text(hjust = 0.5, family = "sans", size = 12),  
        axis.text.x = element_text(color = "black", size = 10),
        axis.text.y = element_text(color = "black", size = 10),
        axis.title.x = element_blank(),
        legend.text = element_text(size = 7),
        legend.title = element_text(size = 7), 
        legend.key.size = unit(0.7, "lines"),
        legend.position =  'bottom',
      #  legend.justification = c(0.2, 1.1),
      #  legend.box.margin = margin(t = 0, r = 0, b = 0, l = 0)
     ) +
  guides(fill = guide_legend(nrow = 2))

ggsave('figures/runtime.pdf', width = 3, height = 3)
```

### Comparison of Index Values

```{r}
library(ggrastr)
# CAI
codonw_cai <- read.table('yeast_cai1.txt', header = T)
cai_comp <- cbind(codonw_cai, data.frame(cai))

ggplot(cai_comp, aes(x = cai, y = CAI)) +
     geom_point_rast(pch = 1, size = 0.5, alpha = 0.4) +
     labs(title = "CAI",
          x = "cubar", y = "CodonW") +
     theme_bw() +
     theme(panel.grid=element_blank(),
           plot.title = element_text(hjust = 0.5, family = "sans", size = 12),  
           axis.text = element_text(color = "black", size = 10))
ggsave('figures/cai_comp.pdf', width = 3, height = 3)

# Fop
codonw_fop <- read.table('yeast_fop1.txt', header = T)
fop_comp <- cbind(codonw_fop, data.frame(fop))

ggplot(fop_comp, aes(x = fop, y = Fop)) +
     geom_point_rast(pch = 1, size = 0.5, alpha = 0.4) +
     labs(title = "Fop",
          x = "cubar", y = "CodonW") +
     theme_bw() +
     theme(panel.grid=element_blank(),
           plot.title = element_text(hjust = 0.5, family = "sans", size = 12),
           axis.text = element_text(color = "black", size = 10)  )
ggsave('figures/fop_comp.pdf', width = 3, height = 3)

# GC
codonw_gc <- read.table('yeast_gc1.txt', header = T)
gc_comp <- cbind(codonw_gc, data.frame(gc))

ggplot(gc_comp, aes(x = gc, y = GC)) +
     geom_point_rast(pch = 1, size = 0.5, alpha = 0.4) +
     labs(title = "GC",
          x = "cubar", y = "CodonW") +
     theme_bw() +
     theme(panel.grid=element_blank(),
           plot.title = element_text(hjust = 0.5, family = "sans", size = 12),
           axis.text = element_text(color = "black", size = 10)  )
ggsave('figures/gc_comp.pdf', width = 3, height = 3)

# GC3s
codonw_gc3s <- read.table('yeast_gc3s1.txt', header = T)
gc3s_comp <- cbind(codonw_gc3s, data.frame(gc3s))

ggplot(gc3s_comp, aes(x = gc3s, y = GC3s)) +
     geom_point_rast(pch = 1, size = 0.5, alpha = 0.4) +
     labs(title = "GC3s",
          x = "cubar", y = "CodonW") +
     theme_bw() +
     theme(panel.grid=element_blank(),
           plot.title = element_text(hjust = 0.5, family = "sans", size = 12),
           axis.text = element_text(color = "black", size = 10)  )
ggsave('figures/gc3s_comp.pdf', width = 3, height = 3)

# RSCU
cubar_rscu_aa <- est_rscu(count_codons(Biostrings::readDNAStringSet(in_fa)), level = 'amino_acid')
codonw_rscu <- read.table('codonw_rscu.txt', col.names = c('codon', 'codonw_rscu'))
codonw_rscu$codon <- gsub('U', 'T', codonw_rscu$codon)
rscu_comp <- left_join(cubar_rscu_aa[,c('codon', 'rscu')], codonw_rscu, by = 'codon')

ggplot(rscu_comp, aes(x = rscu, y = codonw_rscu)) +
     geom_point_rast(pch = 1, size = 0.5, alpha = 0.4) +
     labs(title = "RSCU ",
          x = "cubar", y = "CodonW") +
     theme_bw() +
     theme(panel.grid=element_blank(),
           plot.title = element_text(hjust = 0.5, family = "sans", size = 12),  
           axis.text = element_text(color = "black", size = 10))
ggsave('figures/rscu_comp.pdf', width = 3, height = 3)

# tAI
cf <- count_codons(Biostrings::readDNAStringSet(in_fa))
cubar_w <- est_trna_weight(yeast_trna_gcn, s = list(IC = 0.28, IA = 0.9999, GU = 0.41, UG = 0.68, IU = 0))
cubar_tai <- get_tai(cf, cubar_w[-33,])

codon_table = get_codon_table()
codon_table[, anticodon := as.character(Biostrings::reverseComplement(Biostrings::DNAStringSet(codon_table$codon)))]
codon_table[, ac_level := yeast_trna_gcn[paste(amino_acid, anticodon, sep = '-')]]
codon_table$ac_level[is.na(codon_table$ac_level)] <- 0
tai_w <- tAI::get.ws(tRNA=codon_table$ac_level, sking=0)
tai_tai <- tAI::get.tai(cf[, codon_table$codon[c(-11, -12, -15, -36)]],tai_w)

tai_comp <- cbind(data.frame(tai_tai), data.frame(cubar_tai))
ggplot(tai_comp, aes(x = cubar_tai, y = tai_tai)) +
     geom_point_rast(pch = 1, size = 0.5, alpha = 0.4) +
     labs(title = "tAI",
          x = "cubar", y = "tAI") +
     theme_bw() +
     theme(panel.grid=element_blank(),
           plot.title = element_text(hjust = 0.5, family = "sans", size = 12),
           axis.text = element_text(color = "black", size = 10)  )
ggsave('figures/tai_comp.pdf', width = 3, height = 3)
```

save the ENC index values and copy the content of the output to codonu_enc.txt and dambe_enc.txt
```{python}
# CodonU ENC
an.calculate_enc(in_fa, genetic_code_num = 1, gene_analysis=True, save_file = True)
```

```{r}
# cubar ENC
cubar_enc <- get_enc(count_codons(Biostrings::readDNAStringSet(in_fa)), level = 'subfam')
# CodonW ENC
codonw_enc <- read.table('yeast_enc1.txt', header = T)
#codonw_enc$Nc <- as.numeric(codonw_enc$Nc)
# DAMBE ENC
dambe_enc <- read.table('dambe_enc.txt', header = T)
#dambe_enc <- dambe_enc$Nc
# CodonU ENC
codonu_enc <- read.table('codonu_enc.txt', col.names = c('gene_id', 'CodonU'))

enc_comp <- cbind(as.data.frame(cubar_enc), codonw_enc$Nc, dambe_enc$Nc)
enc_comp$gene_id <- rownames(enc_comp)
enc_comp <- enc_comp %>% left_join(codonu_enc, by = 'gene_id')
colnames(enc_comp)[1:3] <- c('cubar', 'CodonW', 'DAMBE')
enc_comp$CodonW <- as.numeric(enc_comp$CodonW)

scatter_hexbin <- function(data, mapping, ...) {
  ggplot(data = data, mapping = mapping) +
    geom_hex(...) +
    scale_fill_viridis_c()
}

ggpairs(enc_comp[,c(1:3,5)],
        upper = list(continuous = wrap("cor", color = "black")), 
        lower = list(continuous = scatter_hexbin),
       ) +
     theme_bw() +
     theme(panel.grid=element_blank(),
           axis.text = element_text(color = "black", size = 11),
           strip.background = element_blank()  )

ggsave('figures/enc_comp.pdf', width = 8.5, height = 5.5)
```