---
title: "Regression-based framework for codon usage analysis"
format: html
editor_options: 
  chunk_output_type: console
---

show that Generalized model is more flexible for codon usage analysis

- regression of codon usage ~ gene expression (log1p) + mRNA stability (log?)

```{r}
library(cubar)
library(tidyverse)
```

### Show the principles of optimal codon inference
Prepare datasets
```{r}
yeast_cds_qc <- check_cds(yeast_cds)
yeast_cds_qc

yeast_half_life <- as_tibble(yeast_half_life)
yeast_half_life

# merge expression levels and half lives 
yeast_exp2 <- yeast_exp |> 
    full_join(yeast_half_life[c('gene_id', 'half_life')], by = join_by(gene_id)) |> 
    filter(gene_id %in% names(yeast_cds_qc), fpkm >= 1)

yeast_cds_qc <- yeast_cds_qc[names(yeast_cds_qc) %in% yeast_exp2$gene_id]
yeast_cds_qc <- yeast_cds_qc[yeast_exp2$gene_id]
yeast_cf <- count_codons(yeast_cds_qc)

# only keep those with half-life estimated and scale for regression
yeast_exp3 <- yeast_exp2 |> 
    filter(!is.na(half_life)) |> 
    mutate(
        enc = get_enc(yeast_cf[gene_id, ]),
        fpkm2 = scale(log10(fpkm))[, 1],
        half_life2 = scale(log2(half_life))[, 1])
```


```{r}
glm_codon_data <- function(cf, codon_table = get_codon_table(), level = 'subfam',
                      gene_scores, fdr = 0.001){
    cf <- cf[gene_scores$gene_id, ]
    # exclude stop codons
    codon_table <- codon_table[aa_code != '*']
    cf <- cf[, colnames(cf) %in% codon_table$codon]

    # codon data
    lapply(split(codon_table$codon, f = codon_table[[level]]), function(x){
        cf_grp <- cf[, x, drop = FALSE]
        
        total <- rowSums(cf_grp)
        res <- apply(cf_grp, 2, function(x){
            dtt <- data.table::copy(gene_scores)
            dtt$total <- total
            dtt$x <- x
            dtt
        }, simplify = FALSE)
    })
}

yeast_codon_data <- glm_codon_data(yeast_cf, gene_scores = yeast_exp3, level = 'amino_acid')

plt <- bind_rows(yeast_codon_data$Gly, .id = 'codon')

plt |> 
    filter(total >= 10) |> 
    ggplot(aes(x = fpkm, y = x/total)) +
        geom_hex() +
        scale_fill_viridis_c() +
        scale_x_continuous(transform = 'log10') +
        facet_wrap(vars(codon), scales = 'free_y', nrow = 1) +
        theme_bw(base_size = 12) +
        theme(strip.background = element_blank()) +
        labs(x = 'Gene expression level (FPKM)', y = 'Usage among glycine codons')
ggsave('figures/codon_usage_vs_gene_expression_example.pdf', width = 12, height = 3)

# count of amino acids with synonymous codons in each gene
plt2 <- yeast_codon_data |> 
    map(bind_rows, .id = 'codon') |> 
    bind_rows(.id = 'amino_acid') |> 
    distinct(gene_id, gene_name, fpkm, amino_acid, total) |> 
    filter(!amino_acid %in% c('Met', 'Trp'))

ggplot(plt2, aes(x = total, color = amino_acid)) +
    geom_vline(xintercept = 10, linetype = 2, color = 'grey') +
    geom_hline(yintercept = 0.5, linetype = 2, color = 'grey') +
    stat_ecdf() +
    labs(x = 'Number of amino acids', y = 'Cumulative fractions', color = NULL) +
    scale_x_continuous(
        transform = 'log1p',
        breaks = c(0, 1, 2, 5, 10, 20, 50, 100, 200, 500)) +
    scale_y_continuous(expand = expansion(mult = 0.01)) +
    cols4all::scale_color_discrete_c4a_cat('tol.rainbow') +
    theme_bw(base_size = 12) +
    guides(color = guide_legend(ncol = 2))
ggsave('figures/yeast_amino_acid_usage.pdf', width = 5, height = 3)
```


### Compare methods for optimal codon estimation
Analysis
```{r}
est_optimal_corr <- function(
        cf, codon_table = get_codon_table(), level = "subfam",
        gene_score = NULL, fdr = 0.001, method_corr = 'pearson'){
    if (is.null(gene_score)) {
        enc <- get_enc(cf, codon_table = codon_table)
        gene_score <- -enc
    }
    codon_table <- codon_table[aa_code != "*"]
    cf <- cf[, colnames(cf) %in% codon_table$codon]
    binreg <- lapply(split(codon_table$codon, f = codon_table[[level]]), 
        function(x) {
            cf_grp <- cf[, x, drop = FALSE]
            if (ncol(cf_grp) == 1) {
                data.table::data.table(codon = colnames(cf_grp), 
                  coef = 0, pvalue = 1, eff_genes = nrow(cf))
            }
            else {
                total <- rowSums(cf_grp)
                prop_grp <- cf_grp/total
                prop_grp <- prop_grp[total >= 10, ]
                score_grp <- gene_score[total >= 10]
                res <- apply(prop_grp, 2, function(x) {
                  ct <- cor.test(x, score_grp, method = method_corr)
                  c(ct$estimate, ct$p.value, n = length(x))
                })
                res <- data.table::as.data.table(as.data.frame(t(res)), 
                  keep.rownames = "codon")
                data.table::setnames(res, c("codon", "coef", "pvalue", "eff_genes"))
            }
        })
    bingreg <- data.table::rbindlist(binreg, idcol = level)
    bingreg[, `:=`(qvalue, stats::p.adjust(pvalue, method = "BH"))]
    bingreg <- codon_table[bingreg, on = c("codon", level)]
    bingreg[, `:=`(optimal, coef > 0 & qvalue < fdr)]
    return(bingreg)
}

opcodons <- bind_rows(
    enc_pearson = est_optimal_corr(yeast_cf),
    enc_spearman = est_optimal_corr(yeast_cf, method_corr = 'spearman'),
    enc_glm = est_optimal_codons(yeast_cf),
    exp_pearson = est_optimal_corr(yeast_cf, gene_score = log1p(yeast_exp2$fpkm)),
    exp_spearman = est_optimal_corr(yeast_cf, , gene_score = log1p(yeast_exp2$fpkm), method_corr = 'spearman'),
    exp_glm = est_optimal_codons(yeast_cf, gene_score = log1p(yeast_exp2$fpkm)),
    hlf_pearson = est_optimal_corr(yeast_cf[yeast_exp3$gene_id, ], gene_score = yeast_exp3$half_life2),
    hlf_spearman = est_optimal_corr(yeast_cf[yeast_exp3$gene_id, ], gene_score = yeast_exp3$half_life2, method_corr = 'spearman'),
    hlf_glm = est_optimal_codons(yeast_cf[yeast_exp3$gene_id, ], gene_score = yeast_exp3$half_life2),
    .id = 'method')

# Cysteine example: GLM more powerful
opcodons |> 
    filter(str_detect(method, 'hlf'), amino_acid == 'Cys')


# compare coefficients
opcodons_coef <- opcodons |> 
    pivot_wider(id_cols = c(aa_code, codon, amino_acid, subfam),
                names_from = method, values_from = coef) |> 
    rowwise() |> 
    mutate(n_pos = sum(c_across(enc_pearson:exp_glm) > 0),
           n_neg = sum(c_across(enc_pearson:exp_glm) < 0)) |>
    ungroup() |> 
    filter(!amino_acid %in% c('Met', 'Trp'))

opcodons_coef |> 
    select(enc_pearson:exp_glm) |> 
    GGally::ggpairs()

ggplot(opcodons_coef, aes(x = exp_pearson, y = exp_glm)) +
    geom_point() +
    theme_bw(base_size = 12) +
    geom_vline(xintercept = 0, linetype = 2, color = 'grey') +
    geom_hline(yintercept = 0, linetype = 2, color = 'grey') +
    labs(x = 'Correlation coefficient', y = 'GLM coefficient')
ggsave('figures/yeast_exp_glm_coef_vs_pearson_corr.pdf', width = 3, height = 3)

opcodons_pval <- opcodons |> 
    pivot_wider(id_cols = c(aa_code, codon, amino_acid, subfam),
                names_from = method, values_from = pvalue) |> 
    filter(!amino_acid %in% c('Met', 'Trp'))

ggplot(opcodons_pval, aes(x = log10(exp_pearson), y = log10(exp_glm))) +
    geom_point() +
    theme_bw(base_size = 12) +
    geom_abline(slope = 1, intercept = 0, linetype = 2, color = 'grey') +
    labs(x = 'log10(Correlation P value) ', y = 'log10(GLM P value)')
ggsave('figures/yeast_exp_glm_pval_vs_pearson_corr.pdf', width = 3, height = 3)

## other checks 
# compare sign
opcodons_multi <- opcodons |> 
    mutate(gsize = n(), .by = c(subfam, method)) |> 
    filter(gsize > 1)

ggplot(opcodons_multi, aes(x = codon, y = method, fill = factor(sign(coef)))) +
    geom_tile(color = 'black', size = 1) +
    labs(x = NULL, y = NULL, fill = 'Sign(coef)') +
    theme_minimal(base_size = 12) +
    theme(
        axis.text = element_text(color = 'black'),
        axis.text.x = element_text(angle = 90))

# compare optimal codons
opcodons_best <- opcodons |> 
    filter(coef > 0) |> 
    group_by(method, subfam) |> 
    slice_min(qvalue)

opcodons_best_wide <- opcodons_best |> 
    pivot_wider(id_cols = c(aa_code, amino_acid, subfam),
                names_from = method, values_from = codon) |> 
    rowwise() |> 
    mutate(uniq_n = length(unique(c_across(enc_pearson:exp_glm)))) |>
    ungroup()
```

drawback of correlation-based estimation
- coefs for each family were estimated on a different subset genes
- much less genes were used to estimate for rare amino acids, and those genes typically have lower codon bias.

### Multivariate analysis
```{r}
glm_codon <- function(cf, codon_table = get_codon_table(), level = 'subfam',
                      gene_scores, fdr = 0.001){
    cf <- cf[gene_scores$gene_id, ]
    # exclude stop codons
    codon_table <- codon_table[aa_code != '*']
    cf <- cf[, colnames(cf) %in% codon_table$codon]

    # regression analysis for each codon sub-family
    binreg <- lapply(split(codon_table$codon, f = codon_table[[level]]), function(x){
        cf_grp <- cf[, x, drop = FALSE]
        if(ncol(cf_grp) == 1){
            data.table::data.table(
                codon = colnames(cf_grp), feature = c('fpkm2', 'half_life2'),
                coef = 0, se = 0, zvalue = 0, pvalue = 0)
        }else{
            total <- rowSums(cf_grp)
            res <- apply(cf_grp, 2, function(x){
                dtt <- data.table::copy(gene_scores)
                dtt$total <- total
                dtt$x <- x
                fit <- stats::glm(
                    cbind(x, total - x) ~ fpkm2 + half_life2,
                    family = 'binomial', data = dtt)
                data.table::as.data.table(
                    summary(fit)$coefficients[-1, ], keep.rownames = 'feature')
            }, simplify = FALSE)
            res <- data.table::rbindlist(res, idcol = 'codon')
            data.table::setnames(res, c('codon', 'feature', 'coef', 'se', 'zvalue', 'pvalue'))
        }
    })
    bingreg <- data.table::rbindlist(binreg, idcol = level)
    bingreg[, qvalue := stats::p.adjust(pvalue, method = 'BH')]
    bingreg <- codon_table[bingreg, on = c('codon', level)]
    bingreg[, c('se', 'zvalue') := NULL]
    return(bingreg)
}

ggplot(yeast_exp3, aes(x = fpkm, y = half_life)) +
    geom_point() +
    scale_x_log10() +
    scale_y_log10()

ggplot(yeast_exp3, aes(x = fpkm)) + geom_histogram()
ggplot(yeast_exp3, aes(x = fpkm)) + geom_histogram() + scale_x_log10()
ggplot(yeast_exp3, aes(x = half_life)) + geom_histogram()
ggplot(yeast_exp3, aes(x = half_life)) + geom_histogram() + scale_x_log10()

cor.test(yeast_exp3$fpkm, yeast_exp3$half_life, method = 'spearman')
# rho = 0.2863122, correlation is moderate

cub_fit <- glm_codon(yeast_cf, gene_scores = yeast_exp3)
cub_fit <- cub_fit |>
    filter(!amino_acid %in% c('Met', 'Trp')) |> 
    mutate(feature = if_else(feature == 'fpkm2', 'FPKM', 'Half life'),
           label = if_else(qvalue < 0.05, '', 'N.S.'))

ggplot(cub_fit, aes(x = feature, y = codon, fill = coef)) +
    geom_tile(color = 'grey') +
    geom_text(aes(label = label), size = 3) +
    scale_fill_gradient2(high = 'red', low = 'blue') +
    facet_grid(rows = vars(amino_acid), scales = 'free', space = 'free') +
    labs(x = NULL, y = NULL, fill = 'GLM coef') +
    theme_bw(base_size = 12) +
    theme(axis.text = element_text(color = 'black'))
ggsave('figures/cub_glm_fpkm_halflife.pdf', width = 3, height = 12)

```

